---
title: "Final Project"
author: "Sri Sai Nandini Ravi"
date: "2024-12-13"
output: pdf_document
---

```{r}
#Load Libraries
library(readr)
library(readxl)
library(dplyr)
```
```{r, results='hide'}
# Set the file paths
data_file <- "C:/Users/ravis/Desktop/final project big data/NSCH_2022e_Topical_CSV_CAHMI_DRCv2.csv"
labels_file <- "C:/Users/ravis/Desktop/final project big data/NSCH_2022e_CAHMI_DRCv2_CSV_Variable labels.xlsx"

# Load the main dataset
main_data <- read_csv(data_file)

# Load the variable labels
variable_labels <- read_excel(labels_file)

# View the structure of the datasets
glimpse(main_data)
glimpse(variable_labels)

```

```{r, results='hide'}
# Summary statistics for all columns
summary(main_data)

```
Selecting NeuroGenerative Disease related Variables.
```{r}
# Search for NDD-related terms in column names
ndd_keywords <- c("disability", "ADHD", "autism", "learning", "behavior", "Cerebral Palsy","Tourette Syndrome", "delay", "seizure", "speech", "mental", "development")
ndd_associated_vars <- grep(paste(ndd_keywords, collapse = "|"), names(main_data), 
                            value = TRUE, ignore.case = TRUE)

# View the identified variables
ndd_associated_vars

```
The variables selected for this analysis are 
palsy_22 -	Children who currently have cerebral palsy
DownSynd_22 -	Children who have Down Syndrome
seizure_22 -	Children who currently have epilepsy or seizure disorder
tourette_22 -	Children who currently have Tourette Syndrome, age 3-17 years
anxiety_22 -	Children who currently have anxiety problems, age 3-17 years
depress_22 -	Children who currently have depression, age 3-17 years
behavior_22 -	Children who currently have behavioral or conduct problems, age 3-17 years
DevDelay_22 -	Children who currently have developmental delay, age 3-17 years
IntDisab_22 -	Children who currently have intellectual disability or mental retardation, age 3-17 years
speech_22 -	Children who currently have speech or other language disorder, age 3-17 years
learning_22 -	Children who currently have a learning disability, age 3-17 years
autism_22 -	Children who currently have autism or autism spectrum disorder including Asperger's Disorder, pervasive developmental disorder, age 3-17 years
ADHD_22	 - Children who currently have Attention Deficit Disorder (ADD) or Attention-Deficit Hyperactivity Disorder (ADHD), age 3-17 years

```{r}

# List of variables associated with NDD
ndd_variables1 <- c(
  "palsy_22",        # Cerebral palsy
  "DownSynd_22",     # Down Syndrome
  "seizure_22",      # Epilepsy or seizure disorder
  "tourette_22",     # Tourette Syndrome
  "anxiety_22",      # Anxiety problems
  "depress_22",      # Depression
  "behavior_22",     # Behavioral or conduct problems
  "DevDelay_22",     # Developmental delay
  "IntDisab_22",     # Intellectual disability
  "speech_22",       # Speech or language disorder
  "learning_22",     # Learning disability
  "autism_22",       # Autism spectrum disorder
  "ADHD_22"          # ADD or ADHD
)

# Subset the data
ndd_subset1 <- main_data[, ndd_variables1, drop = FALSE]

# View the first few rows of the subsetted data
head(ndd_subset1)

# Save the subsetted dataset to a CSV file
write.csv(ndd_subset1, "C:/Users/ravis/Desktop/final project big data/ndd_subset1.csv", row.names = FALSE)

```
```{r, echo=FALSE}
colSums(is.na(ndd_subset1))
# Replace '99' with NA in the subset
ndd_subset1[ndd_subset1 == 99] <- NA
# Replace all NA values in ndd_subset1 back to 99
ndd_subset1[is.na(ndd_subset1)] <- 99

```
```{r}
summary(ndd_subset1)
```

```{r}
library(ggplot2)
library(gridExtra)

# Loop through all variables and create plots
plots <- list()  # Create an empty list to store plots

for (var in ndd_variables1) {
  # Replace NA with "Missing" and retain original data
  temp_data <- ndd_subset1
  temp_data[[var]] <- ifelse(is.na(temp_data[[var]]), "Missing", as.character(temp_data[[var]]))
  
  # Generate bar plot for each variable
  p <- ggplot(temp_data, aes_string(x = var)) +
    geom_bar(fill = "steelblue") +
    labs(title = paste("Distribution of", var), x = "Severity Level", y = "Count") +
    theme_minimal()
  
  # Store plot in the list
  plots[[var]] <- p
}

# Display all plots (example: print first 4 plots together)
grid.arrange(plots[[1]], plots[[2]], plots[[3]], plots[[4]], nrow = 2)


```
```{r}
# Save the subsetted dataset to a CSV file
write.csv(ndd_subset1, "C:/Users/ravis/Desktop/final project big data/ndd_subset1.csv", row.names = FALSE)
```



```{r, results='hide'}
# View all column names in the dataset
colnames(main_data)

```
Selecting 20 Risk factors or indipendent variables.

```{r}
# Corrected list of selected variables based on actual column names
selected_risk_factors <- c(
  "SC_AGE_YEARS",       # Age
  "SC_SEX",             # Gender
  "SC_RACE_R",          # Race/Ethnicity
  "BMI3_22",            # BMI
  "PHYSACTIV",          # Physical activity
  "SCREENTIME",         # Time spent watching TV or devices
  "HOURSLEEP",          # Sleep hours
  "ALLERGIES",          # Allergies
  "DIABETES",           # Diabetes
  "HEART",              # Heart conditions
  "DiffMem_22",        # difficulty in concentrating 
  "BrainInjTold_22",    # Brain Injury
  "BloodSev_22",       #  Severity of Child's Blood Disorder
  "bullied_22",         # Childern who got bullied
  "GenetSev_22",        # Severity in current genetic or inherited condition
  "smoking_22",         # Smoking environment 
  "NbhdSafe_22",        # Neighborhood safety
  "EmSFamily_22",       # Emotional support from family 
  "MealTogether_22",    # Meals together 
  "housing_22"          # Housing conditions 
)

# Subset the dataset to include only the corrected variables
risk_factor_subset1  <- main_data[, selected_risk_factors, drop = FALSE]

# View the first few rows of the subsetted dataset
head(risk_factor_subset1)



```


```{r, echo=FALSE, results='hide'}
colSums(is.na(risk_factor_subset1))
# Replace '99' with NA in the subset
risk_factor_subset1[risk_factor_subset1 == 99] <- NA
# Replace NA values with 90 or 95 (assumes these were the only changes)
risk_factor_subset1[is.na(risk_factor_subset1)] <- 90  # Replace all NA with 90 (adjust if needed)

summary(risk_factor_subset1)

```

```{r, echo=FALSE, results='hide'}
# Replace unusual values (90, 95) with NA
risk_factor_subset1[risk_factor_subset1 %in% c(90, 95)] <- NA
# Check missing values
colSums(is.na(risk_factor_subset1))
# Replace NA values with 90 or 95 (assumes these were the only changes)
risk_factor_subset1[is.na(risk_factor_subset1)] <- 90  # Replace all NA with 90 (adjust if needed)

# Summarize data
summary(risk_factor_subset1)

```

```{r}
# Impute missing values with the median for numeric columns
risk_factor_subset1 <- risk_factor_subset1 %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))
# Check missing values
colSums(is.na(risk_factor_subset1))
summary(risk_factor_subset1)
```

```{r}
# Save the subsetted data to a new CSV file
write.csv(risk_factor_subset1, "C:/Users/ravis/Desktop/final project big data/risk_factor_subset1.csv", row.names = FALSE)

```

##Now that I have 2 subsets one with NDD associate variables and the other with indipendent variables which are risk factor varaibles, Further we are examining the associations between the characteristics of participants and outcomes.

###Step 1: Explore Associations Between Participant Characteristics and Outcomes

Before we look at relationships, we are examine summary statistics for both subsets:
```{r}

# Summary statistics for NDD subset
summary(ndd_subset1)

# Summary statistics for risk factor subset
summary(risk_factor_subset1)

# Visualize distribution of key variables
library(ggplot2)

# Example: Distribution of Physical Activity Levels
ggplot(risk_factor_subset1, aes(x = factor(DiffMem_22))) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of Physical Activity", x = "Physical Activity Level", y = "Count")

# Distribution of ADHD Severity
ggplot(ndd_subset1, aes(x = factor(ADHD_22))) +
  geom_bar(fill = "lightcoral", color = "black") +
  labs(title = "Distribution of ADHD", x = "ADHD Level", y = "Count")

```

Correlation Analysis
```{r, results='hide'}
# Combine both subsets into a single dataframe
combined_data <- cbind(ndd_subset1, risk_factor_subset1)

# Select only numeric columns for correlation
numeric_data <- combined_data %>% select(where(is.numeric))

# Compute correlation matrix
cor_matrix <- cor(numeric_data, use = "pairwise.complete.obs")

# View the correlation matrix
print(cor_matrix)
```

```{r}
# Heatmap of correlations
library(ggplot2)
library(reshape2)

# Melt correlation matrix for heatmap
melted_cor <- melt(cor_matrix)

# Plot heatmap
ggplot(data = melted_cor, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white",
                       midpoint = 0, limit = c(-1, 1), name = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(title = "Correlation Heatmap", x = "Variables", y = "Variables")

```



```{r}
#Converting the required varaibles to categorical and ordinal to understand the the varaible types. 
# Convert selected variables to categorical (factors)
risk_factor_subset1 <- risk_factor_subset1 %>%
  mutate(
    SC_SEX = as.factor(SC_SEX),
    SC_RACE_R = as.factor(SC_RACE_R),
    PHYSACTIV = as.factor(PHYSACTIV),
    ALLERGIES = as.factor(ALLERGIES),
    DIABETES = as.factor(DIABETES),
    HEART = as.factor(HEART),
    NbhdSafe_22 = as.factor(NbhdSafe_22),
    EmSFamily_22 = as.factor(EmSFamily_22),
    MealTogether_22 = as.factor(MealTogether_22),
    housing_22 = as.factor(housing_22)
  )

# Verify the changes
str(risk_factor_subset1)  # Check data types
summary(risk_factor_subset1)  # Summarize to ensure conversion

```
Statistical Analysis
```{r, results='hide'}
#Chi-Square Tests for each NDD outcome vs risk factor
# Perform Chi-Square Tests for each NDD outcome vs risk factor
chi_square_results <- list()

# Loop over all columns in ndd_subset1
for (outcome in colnames(ndd_subset1)) {
  for (factor in colnames(risk_factor_subset1)) {
    # Chi-Square Test
    test <- chisq.test(table(ndd_subset1[[outcome]], risk_factor_subset1[[factor]]), simulate.p.value = TRUE)
    chi_square_results[[paste(outcome, factor, sep = "_vs_")]] <- test$p.value
  }
}

# Display significant associations
significant_results <- chi_square_results[unlist(chi_square_results) < 0.05]
print("Significant Associations:")
print(significant_results)



```
```{r}
# Example visualization for ADHD and Physical Activity
library(ggplot2)
ggplot(data = cbind(ndd_subset1, risk_factor_subset1), aes(x = PHYSACTIV, fill = factor(ADHD_22))) +
  geom_bar(position = "fill") +
  labs(title = "Association Between Physical Activity and ADHD", x = "Physical Activity", y = "Proportion", fill = "ADHD") +
  theme_minimal()


```
To investigate the relationship between risk factors and outcomes using a machine learning model, using a Random Forest Classifier can handle non-linear relationships, provides insights into variable importance, and works well for datasets with categorical and numerical variables.

```{r, results='hide'}
# Load required libraries
library(randomForest)
library(caret)  # For data splitting and evaluation
# Combine outcomes and risk factors into one dataset
ml_data <- cbind(ndd_subset1, risk_factor_subset1)

# Remove rows with missing values
ml_data <- na.omit(ml_data)

# Verify the structure of the combined dataset
str(ml_data)



```
Split the Data into Training and Testing Sets

```{r}
# Split the data into training (70%) and testing (30%) sets
set.seed(123)  # For reproducibility
trainIndex <- createDataPartition(ml_data$DevDelay_22, p = 0.7, list = FALSE) #  
train_data <- ml_data[trainIndex, ]
test_data <- ml_data[-trainIndex, ]

```

Train the Random Forest Model
Fit the model using the training dataset.
```{r}
# Train a Random Forest model
rf_model <- randomForest(as.factor(DevDelay_22) ~ ., data = train_data, ntree = 500, importance = TRUE)  #
# Print the model summary
print(rf_model)

# Plot variable importance
varImpPlot(rf_model)

```

Evaluate the Model on the Test Set
Assess the model’s performance on the testing dataset.
```{r}
# Predict outcomes on the test set
rf_predictions <- predict(rf_model, test_data, type = "response")

# Confusion matrix
confusionMatrix(as.factor(rf_predictions), as.factor(test_data$DevDelay_22))

```
Analyze Variable Importance
Random Forest provides a ranking of the most important predictors.

```{r}
# Extract variable importance
importance(rf_model)

```

```{r}

# Plot variable importance
varImpPlot(rf_model, main = "Variable Importance")

```
 Investigate Model Performance
We now need to evaluate how well the Random Forest model performed:

Use a Confusion Matrix to check prediction accuracy.
Calculate performance metrics like Accuracy, Precision, Recall, and F1-Score.
```{r}
# Load caret for evaluation metrics
library(caret)

# Predict outcomes using the Random Forest model
predictions <- predict(rf_model, newdata = test_data)

# Convert predictions and true outcomes to factors
predictions <- as.factor(predictions)
true_outcomes <- as.factor(test_data$DevDelay_22) 
# Generate confusion matrix
conf_matrix <- confusionMatrix(predictions, true_outcomes)

# Print confusion matrix and overall statistics
print(conf_matrix)

```
ROC curves are only meaningful for binary classification. If our target variable (e.g., behavior_22) has multiple classes, we need to:

Convert it into a binary outcome (e.g., one class vs. all other classes).

```{r}
# Convert target variable to binary (Class 1 vs. others)
binary_outcome <- ifelse(test_data$DevDelay_22 == 1, 1, 0)
# Predict probabilities for Class 1
predicted_probs <- predict(rf_model, newdata = test_data, type = "prob")[, 1]  # Prob for Class 1
# Load pROC library
library(pROC)

# Create ROC curve
roc_curve <- roc(binary_outcome, predicted_probs)

# Plot the ROC curve
plot(roc_curve, col = "blue", main = "ROC Curve for behavior_22 (Class 1 vs All)")


```
```{r}
# Print the AUC value
auc_value <- auc(roc_curve)
print(paste("AUC:", round(auc_value, 3)))

```


